---
title: "Pre-tutorial instructions"
---

## Softwares

We recommend installing the latest versions of R and Rstudio: 

  * R version 4.4.0, but any version $\geq$ 4.0.0 should work fine,
  * Rstudio version 2024.04.1, but any version $\geq$ 2022.00.00 should work.

An excellent resource for R and Rstudio will be found here: https://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro.html


## Packages

To install packages, you can use `pak`, a great package for easy and comprehensible package installation. First, install `pak` with the usual command

```{r, eval = FALSE}
install.packages("pak")
```

The whole tutorial revolves around the use of five packages: `sGSVD`, `GSVD`, `PTCA4CATA`, `data4PCCAR`, and `SPAFAC`. Install them with `pak`:

```{r, eval = FALSE}
pak::pak(c("vguillemot/sGSVD", "derekbeaton/GSVD", "HerveAbdi/PTCA4CATA", "HerveAbdi/data4PCCAR", "juchiyu/SPAFAC"))
```

Depending on what is already installed on your machine, `pak` will detect what needs to be installed, and give you a summary.


### Other useful packages

> **TL;DR -- here is a one liner to install the packages detailed below.**

```{r install packages, eval = FALSE}
pak::pak(c("TExPosition", "factoextra", "FactoMineR", "tidyverse", "ggrepel", "pheatmap"))
```

  * `TExPosition` for PLS-C, install with `pak::pak("TExPosition")`
  * `FactoMineR` and `factoextra` for PCA, CA, MCA and much much more, install with `pak::pak(c("factoextra", "FactoMineR"))`
  * `tidyverse` for plotting, data wrangling and so much more, install with `pak::pak("tidyverse")`
  * `ggrepel` for adding labels to ggplots, install with `pak::pak("ggrepel")`
  * `pheatmap` for nice heatmaps (but you might prefer `ComplexHeatmap`), install with `pak::pak("pheatmap")`


## Requirements

During the tutorial, we will extensively use `dplyr` and `ggplot2`, we therefore greatly recommend future attendees to refresh their memory on the use of these packages, for example with this course: https://swcarpentry.github.io/r-novice-gapminder


